Draft_Application_Stochastic_Volatility

### 2021-07-19 Real Dataset SV.Rmd
commit

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$. Note that the number of particles was set at 10000 for each filter: the next section will come back to such a choice.} \

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target. 
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.

APF

@@

APF vs Bootstrap

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
library(kableExtra)
kbl(Errorcomp3[,], align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
```

>>>>

We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis. Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows. For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$. Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \

@@

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target. 
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we also conduct this analysis only on one of the three indices, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid degeneracy of the weights on the particles.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.\
As it turns out, such suggestion is confirmed by running the algorithms with N=1000. Passing from 100 to 1000 particles nearly halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a sizable hedge over the Auxiliary PF. Such a hedge almost vanishes when increasing the number of particles to 5000: according to the RMSE, the Bootstrap PF performs slightly better than the Guided PF, which in turn yields very similar results to the Auxiliary PF.\
We would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm. Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (as the RMSE and MAE criteria only prefer the Basic PF to the APF on a couple of occasions), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles. Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.

Now, one might argue that the relative performance of the filters is influenced by the choice of the Bootstrap filter as the benchmark algorithm when run with N=50000. Intuitively, if the convergence phase were not reached, the Bootstrap filter would clearly find it easier to replicate the benchmark we set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.} Such criticism should definitely be taken into account, even more so when noticing that the BT filter did in fact come out on top for several values of N.\
As a robustness check, we thus conduct the same analysis employing the Auxiliary Particle Filter as the benchmark, keeping the number of particles set to 50000 as for the benchmark BT filter.

@@

Using the APF as the benchmark does not deliver a different picture than the one we previously had. Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
Its relative performance ameliorates dramatically at N=5000, when the halving of its RMSE and MAE allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF. At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
Interestingly, this time the performance of the Bootstrap filter worsens when N is increased from 5000 to 10000 (both according to RMSEs and MAEs, whose disagreements, once again, are overall infrequent). Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\
Indeed, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=5000 is closer to the benchmark APF than the benchmark BT is.\footnote{According to the RMSE, also the Bootstrap with N=10000 is closer than the Bootstrap benchmark; however, in this occasion the MAE disagrees. Similarly, while according to the RMSE the APF with N=10000 is closer to the benchmark BT than the benchmark APF is, the MAE is lowest for the benchmark APF.}

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
library(kableExtra)
kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
```

Nevertheless, we should not overplay tiny differences. Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
Second, when the number of particles is set at 5000 or more, neither benchmark could single out an algorithm that clearly outperformed the others, when restricting the analysis to procedures that feature adaptive resampling rules.\
Third, when the number of particles was set to 10, 100 or 1000, the Bootstrap particle filter and the Guided particle filter with optimal proposal kernel significantly emerged as the best performing algorithms for our SV model, suggesting that their computational cost is lower than that of the Auxiliary particle filter, which needed many more particles to "get rolling".\
Finally, some discrepancy between different approximations persisted even when the number of particles was set to 50000. However, the narrow extent of the RMSEs and MAEs with respect to both benchmarks supports the idea that at N=10000 the best-performing algorithms have already entered the phase of convergence. Even in light of these markedly more computationally expensive benchmarks, we then choose N=10000 as the number of particles which we routinely employ in this empirical application.
